---
title: "Post title in sentence case"
description: "One or two sentences. This is what shows on the hub listing and in RSS — it is the only thing most people read, so make it the claim, not a teaser."
date: 2026-01-01
categories:
  - stats-cases
draft: true

# ── Pick exactly ONE engine block. Delete the other. ─────────────────────────
# Do not mix R and Python chunks in one post. reticulate is installed and would
# work, but it doubles the failure surface per post for no benefit here.

# R post — knitr. Chunks are ```{r}.
engine: knitr

# Python post — jupyter. Chunks are ```{python}.
# The kernel is named, NOT pointed at a python.exe. The `ds-general` kernelspec
# carries an env block (PATH including <env>\Library\bin, CONDA_PREFIX) that
# makes it behave as activated. Aiming Quarto at a bare interpreter instead
# makes numpy fail to link libblas.dll: native crash 0xc06d007f, exit 127,
# NO TRACEBACK. See CLAUDE.md § The engine contract.
#
# engine: jupyter
# jupyter: ds-general
# ─────────────────────────────────────────────────────────────────────────────
---

<!--
  HOW TO START A POST

    Copy-Item -Recurse posts\_template posts\<slug>

  `<slug>` becomes the permanent URL and must never change afterwards. No date
  prefix — the `date:` field above drives sort order, so a post can be re-dated
  without breaking its URL. Use lowercase words separated by hyphens, and name
  the subject rather than the conclusion (`logistic-separation`, not
  `why-logistic-regression-broke`), because the conclusion may change.

  Every category slug must already be listed in CATEGORIES.md. Quarto creates a
  new category on a typo without warning.

  `draft: true` above keeps this out of every listing. Delete that line to
  publish. Leave it while drafting — no separate branch needed.

  Assets (data, images) go in this folder alongside index.qmd. Render just this
  post while writing:

    quarto render posts/<slug>/index.qmd

  A full `quarto render` is only needed before publishing.

  Delete this comment block.
-->

Open with the claim. A reader who stops after one paragraph should still come
away with the finding, not just the topic.

## Setup

```{r}
#| label: setup
#| message: false
library(ggplot2)
```

## The thing itself

Label chunks (`#| label:`) — the label names the freeze artifact and the figure
anchor, so labelled chunks produce stable cross-references and readable diffs.

```{r}
#| label: fig-example
#| fig-cap: "Captions get a figure number and become linkable as @fig-example."
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  theme_minimal()
```

## What this does not show

State the limits explicitly. The simulations here are only as good as their
assumptions, and saying which ones are load-bearing is most of the value.

## Reproducing this

This post's executed output is frozen and committed. To re-run it from source
rather than replaying the cache:

```powershell
quarto render posts/<slug>/index.qmd --no-freeze
```

If that produces different numbers than what is shown above, that is the
trigger to pin dependencies with `renv` — see CLAUDE.md § Deferred.
