---
title: "Separation makes logistic regression fail loudly in the warnings and quietly in the estimates"
description: "As a predictor gets closer to perfectly separating a binary outcome, the maximum likelihood estimate diverges. R warns about fitted probabilities, not about the coefficient — which is the number people actually report."
date: 2026-08-12
categories:
  - stats-cases
engine: knitr
---

When a predictor separates a binary outcome perfectly, the logistic regression
maximum likelihood estimate does not exist. The likelihood keeps increasing as
the coefficient grows, so the optimiser stops at whatever its convergence rule
happens to hit. The reported number is an artefact of the stopping rule.

R does warn, but it warns about the wrong thing: `fitted probabilities
numerically 0 or 1 occurred` describes a symptom on the probability scale. The
coefficient and its standard error — the two numbers that end up in a table —
come back with no annotation at all.

This post simulates the approach to separation so you can watch both quantities
go.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false
library(ggplot2)
library(ggiraph)
library(plotly)

set.seed(20260812)
```

## Simulating the approach to separation

One predictor, one binary outcome. The parameter `beta` controls how sharply
the outcome tracks the predictor: at `beta = 0.5` the classes overlap heavily,
and as `beta` grows the boundary sharpens until — in any finite sample — the
two classes stop overlapping at all. That last state is separation, and it
arrives by degrees rather than all at once.

```{r}
#| label: simulate
simulate_once <- function(beta, n = 200) {
  x <- rnorm(n)
  y <- rbinom(n, size = 1, prob = plogis(beta * x))
  data.frame(x = x, y = y)
}

# `glm` warns about fitted probabilities near separation; that warning is the
# subject of this post rather than a problem with the call, so it is captured
# rather than printed 40 times.
fit_once <- function(beta) {
  d <- simulate_once(beta)
  warned <- FALSE
  m <- withCallingHandlers(
    glm(y ~ x, data = d, family = binomial),
    warning = function(w) {
      warned <<- TRUE
      invokeRestart("muffleWarning")
    }
  )
  co <- summary(m)$coefficients
  overlap <- min(max(d$x[d$y == 0]), max(d$x[d$y == 1])) -
    max(min(d$x[d$y == 0]), min(d$x[d$y == 1]))

  data.frame(
    beta_true = beta,
    estimate  = co["x", "Estimate"],
    se        = co["x", "Std. Error"],
    warned    = warned,
    overlap   = overlap
  )
}

betas <- seq(0.5, 18, length.out = 45)
res <- do.call(rbind, lapply(betas, fit_once))
```

## The estimate

Hover any point for the fitted values. Points where R issued a warning are
marked.

```{r}
#| label: fig-estimates
#| fig-cap: "Estimated coefficient against its true value. The diagonal is where an unbiased estimator would sit."
res$tooltip <- sprintf(
  "true beta: %.1f\nestimate: %.2f\nstd. error: %.2f\nR warned: %s",
  res$beta_true, res$estimate, res$se, ifelse(res$warned, "yes", "no")
)

p <- ggplot(res, aes(beta_true, estimate)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey55") +
  geom_line(colour = "grey70") +
  geom_point_interactive(
    aes(tooltip = tooltip, data_id = beta_true, colour = warned),
    size = 2.8
  ) +
  scale_colour_manual(
    values = c("FALSE" = "#0f766e", "TRUE" = "#b45309"),
    labels = c("FALSE" = "no warning", "TRUE" = "R warned"),
    name = NULL
  ) +
  labs(x = "true coefficient", y = "estimated coefficient") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")

girafe(
  ggobj = p,
  width_svg = 7,
  height_svg = 4.2,
  options = list(
    opts_hover(css = "fill:#c026d3;stroke:none;r:5pt;"),
    opts_tooltip(css = "background:#111;color:#eee;padding:6px;border-radius:4px;font-family:monospace;font-size:12px;")
  )
)
```

The estimate tracks the truth while the classes still overlap and then comes
apart. Past the point where a sample happens to be separable, the estimate is
reporting the optimiser's patience, not the data.

## The standard error

The standard error is the more dangerous of the two, because it moves in the
direction that looks like caution:

```{r}
#| label: fig-se
#| fig-cap: "Standard error of the coefficient. Note the log scale — this is not a gentle increase."
q <- ggplot(res, aes(beta_true, se, colour = warned)) +
  geom_line(colour = "grey70") +
  geom_point(size = 2) +
  scale_y_log10() +
  scale_colour_manual(
    values = c("FALSE" = "#0f766e", "TRUE" = "#b45309"),
    guide = "none"
  ) +
  labs(x = "true coefficient", y = "standard error (log scale)") +
  theme_minimal(base_size = 13)

ggplotly(q, tooltip = c("x", "y"))
```

A huge standard error produces a wide confidence interval and a p-value nowhere
near significance. Read casually, that says "not enough evidence" — the exact
opposite of what is happening, which is that the predictor is so strongly
associated with the outcome that the model cannot express it.

## Where the warning first appears

```{r}
#| label: warning-onset
first_warned <- res[res$warned, ][1, ]
knitr::kable(
  first_warned[, c("beta_true", "estimate", "se", "overlap")],
  digits = 2,
  col.names = c("true coefficient", "estimate", "std. error", "class overlap"),
  caption = "The first simulated fit at which R emitted a warning."
)
```

## What this does not show

This is one predictor, one sample per setting, and no repetition — so the
wiggle between adjacent points is sampling noise, not structure. The
qualitative behaviour (divergence, exploding standard error) is a property of
the likelihood and does not depend on the seed; the specific value of `beta`
where it starts does, because whether a given sample is separable is itself
random.

The fix is not shown here. Penalised likelihood (Firth) and a weakly
informative prior both give finite estimates under separation, and comparing
them is its own post.

## Reproducing this

This post's executed output is frozen and committed, so rendering the site
replays it rather than re-running it. To actually re-execute:

```powershell
quarto render posts/logistic-separation/index.qmd --no-freeze
```

Different numbers from a re-run would mean a package changed underneath the
post — that is the trigger to pin dependencies.
